Improving the Stability of Type Soundness Proofs in Dafny

### 2023-10-17 paper/paper.tex
up to date

\begin{abstract}
  In this extended abstract, we present a systematic technique for
  structuring type soundness proofs in Dafny to improve proof stability. As a case
  study, we apply the method to proving type soundness for a small expression
  language, and demonstrate empirically how it improves resource usage metrics
  known to correlate with stability. Our method can scale to realistic proofs,
  as demonstrated by its use in the type soundness proof of the Cedar Policy
  Language.
\end{abstract}

@@

\subsection*{Results}
The graph in Figure~\ref{fig:results} shows the mean verification \emph{resource usage} (and standard deviation, with whiskers), over the 250 verifications of each soundness theorem.

% \begin{itemize}
%   \item We look at the mean verification \emph{duration} and \emph{resource usage}
%         over 250 verifications --- lower is better in both cases, meaning that the proof takes less time and is less costly on average.
%   \item Of greater interest is the \emph{variance} of duration and resource usage. A proof which vareis wildly in its verification duration or resource usage
%         is often a precursor to a proof whose outcome varies. \jwc{Cite Aaron Tomb, personal communication} This is shown with whiskers (standard deviations)
% \end{itemize}

\begin{figure}
  \includegraphics{costgraph.png}
  \caption{Verification Resource Usages}
\end{figure}
\jwc{Graph 1}

The highest cost and highest variance proof in all cases is the soundness proof for the typechecker in \texttt{monadic.dfy}. It strikes an unforuntate balance of being complicated code --- heavy use of monadic bind and lots of helper functions --- with little guidance in the proof.
The typechecker in \texttt{direct.dfy} is slightly cheaper to verify than the one in \texttt{monadic.dfy}, but it still has enormous variance. The cost seems to be lower because of all of the inlined definitions,
meaning that the verifier must reason about fewer functions while proving soundness.
The proof for the typechecker in \texttt{bidirectional.dfy} is cheaper and has lower variance than either of the previous two. This may be because while the bidirectionality increases the complexity of the typechecking algorithm,
this complexity \emph{requires} further guidance to the verifier for it to accept the proof at all, thereby improving stability on the whole.
As expected, the two proofs using the first and second halves of our technique
are by far the best. The soundness theorem using the fully specified typechecker and safety property (\texttt{semantic-invert.dfy})
is the best of all, with negligable variance in resource usage between runs of the verifier.

@@

\subsubsection*{Scaling It Up}
A version of this proof technique was developed for the purposes of mechanizing the proof of type soundness
of the Cedar Policy Language \jwc{cite} \footnote{https://github.com/cedar-policy/cedar-spec/tree/main/cedar-dafny},
and the technique described in this abstract is currently used in the type soundness proof of the language.
Cedar's type system is a great deal more complex than that of our toy langauge, and includes a number of advanced type system features like ocurrence typing \jwc{cite: The Design and Implementation of Typed Scheme} and
singleton types \cite{https://dl.acm.org/doi/10.5555/645867.670933}.

Semantically, Cedar is not \emph{all} that much more complex to model than the
toy language we evaluate for this extended abstract. Like our toy language,
Cedar is terminating and first order.  Although we have not tested it, our
technique should scale to languages with nontermination --- by step-indexing the
evaluator and proving type-safety with failure to converge in $n$ steps being
considered ``safe'' \jwc{cite step-indexing} --- and higher-order functions ---
by applying standard tricks to handle variable binding. \jwc{cite}

@@

\subsection*{Cost of Compatibility and Inverson Lemmas}
\jwc{Decreases instability, but does require extra lemmas! Because the other proofs dont' have these lemmas broken out, we have no way to empircally measure how they compare to the proofs in the other techniques. That being said, we have anecdotatlly found the compatability and inversion lemmas to be very stable with minimal guidance, both in this development and in Cedar.}

\jwc{The compataiblity and inversion lemmas are small and only require reasoning about either the evaluator or the checker.}

>>>>

\begin{abstract}
  In this extended abstract, we present a systematic technique for
  structuring type soundness proofs in Dafny to improve proof stability. As a case
  study, we apply the method to proving type soundness for a small expression
  language, and demonstrate empirically how it improves resource usage metrics
  known to correlate with stability. Our method can scale to realistic proofs,
  as demonstrated by its use in the type soundness proof of the Cedar
  Language.
\end{abstract}

@@

\subsection*{Results}
The graph in Figure~\ref{fig:cost-results} shows the mean verification \emph{resource usage} (and standard deviation, with whiskers), over the 250 verifications of each soundness theorem,
while the graph in Figure~\ref{fig:dur-results} shows verification \emph{durations}, in miliseconds. The two graphs tell essentially the same story.
\jwc{Re-Run the experiments!}
The highest cost and highest variance proof in all cases is the soundness proof for the typechecker in \texttt{monadic.dfy}. It strikes an unforuntate balance of being complicated code --- heavy use of monadic bind and lots of helper functions --- with little guidance in the proof.
The typechecker in \texttt{direct.dfy} is slightly cheaper to verify than the one in \texttt{monadic.dfy}, but it still has enormous variance. The cost seems to be lower because of all of the inlined definitions,
meaning that the verifier must reason about fewer functions while proving soundness.
The proof for the typechecker in \texttt{bidirectional.dfy} is cheaper and has lower variance than either of the previous two. This may be because while the bidirectionality increases the complexity of the typechecking algorithm,
this complexity \emph{requires} further guidance to the verifier for it to accept the proof at all, thereby improving stability on the whole.
As expected, the two proofs using the first and second halves of our technique
are by far the best. The soundness theorem using the fully specified typechecker and safety property (\texttt{semantic-invert.dfy})
is the best of all, with negligable variance in resource usage between runs of the verifier.

\begin{figure}
  \includegraphics[scale=0.5]{durgraph.png}
  \caption{Verification Duration}
  \label{fig:dur-results}
\end{figure}

\begin{figure}
  \includegraphics[scale=0.5]{costgraph.png}
  \caption{Verification Resource Usages}
  \label{fig:cost-results}
\end{figure}

@@

\subsubsection*{Scaling It Up}
A version of this proof technique was developed for the purposes of mechanizing the proof of type soundness
of the Cedar Language \jwc{cite} \footnote{https://github.com/cedar-policy/cedar-spec/tree/main/cedar-dafny},
the language underlying the Amazon Verified Permissions offering \footnote{https://aws.amazon.com/verified-permissions/}.
The technique described in this abstract is currently used in the type soundness proof of the reference typechecker with respect to the reference interpreter.
In Cedar, programs express access control policies, and the evaluator outputs an authorization decision: permit or deny.
Cedar's type system is a great deal more complex than that of our toy langauge, and includes a number of advanced type system features like ocurrence typing \jwc{cite: The Design and Implementation of Typed Scheme} and
singleton types \cite{https://dl.acm.org/doi/10.5555/645867.670933}.

Semantically though, Cedar is not \emph{all} that much more complex to model than the
toy language we evaluate for this extended abstract. Like our toy language,
Cedar's evaluator is simplified by the fact that it is terminating, first order, and has no binders.  Although we have not tested it, our
technique should scale to languages with nontermination --- by step-indexing the
evaluator and proving type-safety with failure to converge in $n$ steps being
considered ``safe'' \jwc{cite step-indexing} --- and higher-order functions ---
by applying standard tricks to handle variable binding. \jwc{cite}

@@
